---
title: "GUIDE IBD"
author: "Joana P. Bernardes and Charlot Jaeckel"
date: "3/4/2021"
output: pdf_document
    fig_width: 7
    fig_height: 6
    fig_caption: true

---

## R Markdown

This document will create the plots for GUIDE-IBD molecular medicine board (MMB) reports.
Each report will be focused on an individual samples. The reference samples were collect from eMed and SYSCID-sc cohorts. In total, we will use 26 genes from blood samples and 16 genes for biopsy samples to tested for early remission patterns by comparing the test samples with the reference samples.
Samples were collected from blood or biopsy tissues enrolled in the MMB setup. These patients are therapy-naive CD and UC patients from Kiel, Hamburg or Hannover.
For more information about the patients please see https://survey.ikmb.uni-kiel.de/en/landing_page/about_us


# Load packages
```{r setup, include=FALSE}
library(ggplot2)
library(gplots)
library(RColorBrewer)
library(plyr)
library(dplyr)
library(reshape2)
```


# set working window
```{r}
require("knitr")
opts_knit$set(root.dir = "~/PatientX/")
setwd( "~/.")
```




# Blood
## load query data (saved as .csv)
```{r}
data<-read.csv2('~/PatientX/table.csv')
data$CT<-as.numeric(sub(",", ".", data$CT, fixed = TRUE))
```

## Data wrangling > From raw CT values to relative normalized values.

```{r}
table(data$Sample.Name)
data_sample<-subset(data, data$Sample.Name %in% 'Blood Sample PatientX')
data_sample = transform(data_sample, Target.Name = colsplit(Target.Name,pattern = "_", names = c('Tissue', 'Assay', 'Gene.ID','Primer' , "Primer2")))

average<-aggregate(x = data_sample$CT,                # Specify data column
          by = list(data_sample$Target.Name$Gene.ID),              # Specify group indicator
          FUN = mean) 
rownames(average)<-average$Group.1
average$Group.1=NULL
average$raw<-average$x
average$x=NULL

# housekeeping
hk1<-subset(average, rownames(average) %in% c('TBP', 'ACTB'))
hk2<-as.data.frame(colMeans(hk1))
norm_hk<-subset(hk2, rownames(hk2) %in% c('raw'))

# biomarkers
genes<- subset(average, rownames(average) %in% c('BUB1B','HIVEP2','SAMD9L','IFI16','CASP5','IFI44', 'IFIT5', 'IL18RAP','CXXC5','KIAA0368', 'TREM1','FAM134B','MKI67','APOL6','FAM208B','ARHGAP35','FCGR1B;FCGR1A;FCGR1CP', 'FCRL5', 'HP','ADGRG1','COMMD3,COMMD3-BMI1','RPRD2','SENP7', 'ZNF839','GBP5', 'GZMK'))
table<-genes
# relative
table$relative<-genes$raw/norm_hk$`colMeans(hk1)`
# normalize
table$relative_norm<-2^-table$relative
table$Test<-table$relative_norm
w14<-table

write.csv(table, file='NormData_PatientX.csv')
```

# load reference data
```{r}
# blood
reference<-read.csv2('~/Reference_Blood.csv', row.names = 1, dec=',')
metadata<-read.csv2('~/Metadata_Blood.csv',row.names = 1)
```

## Merge query with reference data
```{r}
reference$GeneID<-rownames(reference)
table$GeneID<- rownames(table)
w0$GeneID<- rownames(w0)
w2$GeneID<- rownames(w2)
w6$GeneID<- rownames(w6)
w14$GeneID <- rownames(w14)
w26$GeneID <- rownames(w26)
w0<-w0[ ,4:5]
w2<-w2[ ,4:5]
w6<-w6[ ,4:5]
w14<-w14[ ,4:5]
w26<-w26[ ,4:5]


Reference_Test <- merge(reference, w0,
              by = 'GeneID')
Reference_Test2 <- merge(Reference_Test, w2,
              by = 'GeneID')
Reference_Test3 <- merge(Reference_Test2, w6,
              by = 'GeneID')
Reference_Test4 <- merge(Reference_Test3, w14,
              by = 'GeneID')
Reference_Test5 <- merge(Reference_Test4, w26,
              by = 'GeneID')
rownames(Reference_Test4)<-Reference_Test4$GeneID
Reference_Test4$GeneID=NULL


metadata$Test<-'w0'
metadata$Test2<-'w2'
metadata$Test3<-'w6'
metadata$Test4<-'w14'
metadata$Test5<-'w26'
metadata2<-as.data.frame(t(metadata))

# if there are missing values remove before principal component
#genes_to_remove<-c("CXXC5")
Reference_Test5<-Reference_Test4[!(row.names(Reference_Test4) %in% genes_to_remove),]


## PCA for emed and scRNAseq cohorts blood 
pc <-prcomp(na.omit(t(Reference_Test5)))
pc_df <- as.data.frame(pc$x)
pc_df$Type<-rownames(pc_df)
pc_df$Remission<-metadata2$Remission
pc_df$Remission2<-metadata2$Remission2
pc_df$Disease<-metadata2$Disease
pc_df$cohort<-metadata2$cohort

pc_df2<-subset(pc_df, !pc_df$Type %in% c('Test.x', 'Test.y'))
pc_df2<-subset(pc_df, !pc_df$Type %in% c('Test.x', 'Test.y', 'Test.x.1','Test.y.1'))
pc_df2<-subset(pc_df, !pc_df$Type %in% c('Test.x', 'Test.y', 'Test.x','Test.y', 'Test'))
pc_df3<-pc_df[c(95), ]
pc_df4<-pc_df[c(96), ]
pc_df5<-pc_df[c(97), ]
pc_df6<-pc_df[c(98), ]
pc_df7<-pc_df[c(99), ]


```

## Plot MMB PCA
```{r, fig.width = 6, fig.height=5}
eigs <- pc$sdev^2
eigs[1] / sum(eigs)
eigs[2] / sum(eigs)

#run this
# overlay for multiple samples
overlay.fs1 <- pc_df3
overlay.fs2 <- pc_df2
overlay.fs3 <- pc_df4
overlay.fs4 <- pc_df5
overlay.fs5 <- pc_df6
overlay.fs6 <- pc_df7


p<-ggplot(pc_df, aes(x = PC1, y =  PC2,  label=Type, color=Remission2,  group=Remission2)) + 
    geom_point(data = overlay.fs2,size=3)  +  
  xlab("PC1 (47.71% variance)") +  ylab("PC2 (16.22% variance)")+
  theme(text = element_text(size=rel(16))) + theme_bw()+
  #scale_color_manual(values = c( "grey","darkorange1","darkolivegreen3", 'black', 'red', 'blue'))+
  scale_color_manual(values = c( "grey","darkorange1","darkolivegreen3", 'black','forestgreen', 'red', 'blue'))+
  #scale_color_manual(values = c( "grey","darkorange1","darkolivegreen3", 'black','forestgreen', 'red', 'gold', 'blue', 'black'))+
  scale_size_area(max_size=2)+
  stat_ellipse(type = "t", linetype = 3, level = 0.90) +
  stat_ellipse(type = "t", linetype = 2, level = 0.80) +
  stat_ellipse(type = "t",linetype = 1, level = 0.70)
p1 <- p + geom_point(data = overlay.fs1, size=5, color=c('black'))+
  geom_text(data = overlay.fs1, aes(x = PC1, y = PC2, label = "0w", vjust=2))
p2 <- p1 + geom_point(data = overlay.fs3, size=5, color=c('red'))+
  geom_text(data = overlay.fs3, aes(x = PC1, y = PC2, label = "2w", vjust=2))
p3 <- p2 + geom_point(data = overlay.fs4, size=5, color=c('blue'))+
  geom_text(data = overlay.fs4, aes(x = PC1, y = PC2, label = "6w", vjust=2))
p4 <- p3 + geom_point(data = overlay.fs5, size=5, color=c('forestgreen'))+
  geom_text(data = overlay.fs5, aes(x = PC1, y = PC2, label = "14w", vjust=2))
p5 <- p4 + geom_point(data = overlay.fs6, size=5, color=c('gold'))+
  geom_text(data = overlay.fs6, aes(x = PC1, y = PC2, label = "26w", vjust=2))
pdf("PCA_PatientX_Blood1_Blood2_Blood3_Blood4_PC12.pdf", width = 4, height = 3.2)
p4
dev.off()
```


# Biopsy
# load query data (saved as .csv)

## Data wrangling > From raw CT values to relative normalized values.

```{r}
table(data$Sample.Name)

data_sample<-subset(data, data$Sample.Name %in% 'Biopsy Sample PatientX')
data_sample<- transform(data_sample, Target.Name = colsplit(Target.Name,pattern = "_", names = c('Tissue', 'Assay', 'Gene.ID','Primer' , "Primer2")))

average<-aggregate(x = data_sample$CT,                # Specify data column
          by = list(data_sample$Target.Name$Gene.ID),              # Specify group indicator
          FUN = mean) 
rownames(average)<-average$Group.1
average$Group.1=NULL
average$raw<-average$x
average$x=NULL


# housekeeping
hk1<-subset(average, rownames(average) %in% c('TBP', 'ACTB'))
hk2<-as.data.frame(colMeans(hk1))
norm_hk<-subset(hk2, rownames(hk2) %in% c('raw'))

# biomarkers
table<- subset(average, rownames(average) %in% c("AQP9", "BANK1","CHI3L2","CMTM2", "CSF3R", "CXCL13",
                                                 "FAM129C", "FCRL1", "IL13RA2", "MS4A1", "PAX5",
                                                 "PTGS2", "S100A9", "STC1", "TNF", "TREM1" ))

# relative
table$relative<-table$raw/norm_hk$`colMeans(hk1)`
# normalize
table$relative_norm<-2^-table$relative
table$Test<-table$relative_norm
w14<-table

write.csv(table, file='NormData_PatientX_Biopsy_w14.csv')
```

# load reference data
```{r}
# biopsy
reference<-read.csv('~/Documents/IKMB/GUIDE IBD/GUIDEIBD_MMB/Reference_Biopsy.csv', row.names = 1, dec='.')
metadata<-read.csv('~/Documents/IKMB/GUIDE IBD/GUIDEIBD_MMB/Metadata_Biopsy.csv',row.names = 1)
```

## Merge query with reference data
```{r}
reference$GeneID<-rownames(reference)
w0$GeneID<- rownames(w0)
w14$GeneID<- rownames(w14)
w26$GeneID<- rownames(w26)

w0<-w0[ ,4:5]
w14<-w14[ ,4:5]
w26<-w26[ ,4:5]

Reference_Test <- merge(reference, w0,
              by = 'GeneID')
Reference_Test2 <- merge(Reference_Test, w14,
              by = 'GeneID')
Reference_Test3 <- merge(Reference_Test2, w26,
              by = 'GeneID')
rownames(Reference_Test2)<-Reference_Test2$GeneID
Reference_Test2$GeneID=NULL



metadata$Test<-'w0'
metadata$Test2<-'w14'
metadata$Test3<-'w26'
metadata2<-as.data.frame(t(metadata))

# if there are missing values remove before principal component
#genes_to_remove<-c("FAM129C", 'IL13RA2', 'MS4A1')
#Reference_Test3<-Reference_Test2[!(row.names(Reference_Test2) %in% genes_to_remove),]


## PCA for emed and scRNAseq cohorts blood 
pc <-prcomp(na.omit(t(Reference_Test3)))
pc_df <- as.data.frame(pc$x)
pc_df$Type<-rownames(pc_df)
pc_df$Remission<-metadata2$Remission
pc_df$Remission2<-metadata2$Remission2
pc_df$Disease<-metadata2$Disease
pc_df$cohort<-metadata2$cohort

pc_df2<-subset(pc_df, !pc_df$Type %in% c('Test'))
pc_df2<-subset(pc_df, !pc_df$Type %in% c('Test.x', 'Test.y'))
pc_df2<-subset(pc_df, !pc_df$Type %in% c('Test.x', 'Test.y', 'Test'))

pc_df3<-pc_df[c(32), ]
pc_df4<-pc_df[c(33), ]
pc_df5<-pc_df[c(34), ]

```

## Plot MMB PCA
```{r, fig.width = 2.8, fig.height=2}
eigs <- pc$sdev^2
eigs[1] / sum(eigs)
eigs[2] / sum(eigs)


# overlay for multiple samples
overlay.fs1 <- pc_df3
overlay.fs2 <- pc_df2
overlay.fs3 <- pc_df4
overlay.fs4 <- pc_df5

p<-ggplot(pc_df, aes(x = PC1, y =  PC2,  label=Type, color=Remission2,  group=Remission2)) + 
    geom_point(data = overlay.fs2,size=3)  +  
  xlab("PC1 (80.47% variance)") +  ylab("PC2 (13.57% variance)") + 
  theme(text = element_text(size=rel(14))) + theme_bw()+
  #scale_color_manual(values = c( "grey","darkorange1","darkolivegreen3", 'black', 'red', 'blue'))+
  scale_color_manual(values = c( "grey","darkorange1","darkolivegreen3", 'black', 'forestgreen'))+
  #scale_color_manual(values = c( "grey","darkorange1","darkolivegreen3", 'black','forestgreen', 'gold'))+
  scale_size_area(max_size=2)+
  stat_ellipse(type = "t", linetype = 3, level = 0.90) +
  stat_ellipse(type = "t", linetype = 2, level = 0.80) +
  stat_ellipse(type = "t",linetype = 1, level = 0.70)
p1 <- p + geom_point(data = overlay.fs1, size=5, color=c('black'))+
  geom_text(data = overlay.fs1, aes(x = PC1 , y = PC2, label = "0w", vjust=2))
p2 <- p1 + geom_point(data = overlay.fs3, size=5, color=c('forestgreen'))+
  geom_text(data = overlay.fs3, aes(x = PC1, y = PC2, label = "14w", vjust=2))
p3 <- p2 + geom_point(data = overlay.fs4, size=5, color=c('gold'))+
  geom_text(data = overlay.fs4, aes(x = PC1, y = PC2, label = "26w", vjust=2))
pdf("PCA_PatientX_Biopsy1_Biopsy2_PC12.pdf", width = 4, height = 3.2)
p2
dev.off()

```



## Plot MMB heatmap for inflammation score
# add scores from https://survey.ikmb.uni-kiel.de/en/landing_page/about_us
```{r}
scores<-read.csv2('~/Inflammatory score.csv',  dec = ',')
min(scores$CRP, na.rm = TRUE)
max(scores$CRP, na.rm = TRUE)

min(scores$Calprotectin, na.rm = TRUE)
max(scores$Calprotectin, na.rm = TRUE)

min(scores$Leukocyte, na.rm = TRUE)
max(scores$Leukocyte, na.rm = TRUE)

scores$CRP_rel = scores$CRP *100 / 75.7
scores$Calprotectin_rel = scores$Calprotectin *100 / 3916
scores$Leukocyte_rel = scores$Leukocyte *100 / 21.86

scores$name<-paste(scores$Cohort, scores$Patient, scores$TimePoint, scores$Remission, sep = "_")

vars<-c('CRP_rel', 'Calprotectin_rel', 'Leukocyte_rel')
inflamm<-scores[vars]
inflamm$Sum<-rowMeans(inflamm, na.rm=TRUE)
rownames(inflamm)<- scores$name

```

### Go to portal and calculate the inflammation score based on sample (Patient and Timepoint)
```{r}
CRP<-15.3
Calprotectin<-73.2
Leukocyte<-8.6
Patient.w0<-as.data.frame(cbind(CRP, Calprotectin,Leukocyte))

CRP<-4.4
Calprotectin<-104
Leukocyte<-7.5
Patient.w2<-as.data.frame(cbind(CRP, Calprotectin, Leukocyte))

CRP<-6.2
Calprotectin<-NA
Leukocyte<-7.5
Patient.w6<-as.data.frame(cbind(CRP, Calprotectin, Leukocyte))

CRP<-3.4
Calprotectin<-13.60
Leukocyte<-6.1
Patient.w14<-as.data.frame(cbind(CRP, Calprotectin, Leukocyte))

CRP<-4
Calprotectin<-16.7
Leukocyte<-5.7
Patient.w26<-as.data.frame(cbind(CRP, Calprotectin, Leukocyte))

Patient.w0$Time<-'w0'
Patient.w2$Time<-'w2'
Patient.w6$Time<-'w6'
Patient.w14$Time<-'w14'
Patient.w26$Time<-'w26'

Patient<-rbind(Patient.w0, Patient.w2)
Patient<-rbind(Patient.w0, Patient.w2, Patient.w6)
Patient<-rbind(Patient.w0, Patient.w2, Patient.w6, Patient.w14)
Patient<-rbind(Patient.w0, Patient.w2, Patient.w6, Patient.w14, Patient.w26)

rownames(Patient)<-Patient$Time

Patient$CRP_rel = Patient$CRP *100 / 75.7
Patient$Calprotectin_rel = Patient$Calprotectin *100 / 3916
Patient$Leukocyte_rel = Patient$Leukocyte *100 / 21.86

Patient2<-Patient[ ,5:7]
Patient2$Sum<-rowMeans(Patient2, na.rm=TRUE)

lines <- data.frame(vlines =Patient2$Sum, labels = c("0 w", "2 w"), stringsAsFactors = FALSE)
lines <- data.frame(vlines =Patient2$Sum, labels = c("0 w", "2 w", '6 w'), stringsAsFactors = FALSE)
lines <- data.frame(vlines =Patient2$Sum, labels = c("0 w", "2 w", '6 w', '14 w'), stringsAsFactors = FALSE)
lines <- data.frame(vlines =Patient2$Sum, labels = c("0 w", "2 w", '6 w', '14 w', '26 w'), stringsAsFactors = FALSE)

```

### Plot MMB histogram
```{r, fig.width = 3, fig.height=1.8}
histo<-ggplot(inflamm, aes(Sum, fill = cut(Sum, 20) )) +
    geom_histogram(colour="black", binwidth = 2 )+
  scale_fill_discrete(h = c(120, 20), c = 80, l = 60 ) +
  theme_bw() +
  labs(x = "Inflamation score", y = "N samples") +
  theme(legend.position = "none") +
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(n.breaks = 6) +
  geom_vline(data = lines, aes(xintercept = vlines))+
  geom_text(data = lines, aes(x = vlines, y = 12, angle=90,label = labels, hjust=0.8,  vjust=1),
    inherit.aes = FALSE, color=c( 'black','forestgreen', 'red', 'blue'))
   # inherit.aes = FALSE, color=c( 'black','forestgreen', 'red', 'blue', 'gold'))
   # inherit.aes = FALSE, color=c( 'black', 'red'))


pdf("Histogram_PatientX_0w2w6w14w.pdf", width = 5, height = 2.8)
histo
dev.off()

```

